Using the internal language of toposes in algebraic geometry

### 2015-11-30 slides-ihes2015.tex
More note-slides

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{mathtools}
\usepackage{soul}\setul{0.2ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Hom}{\mathrm{Hom}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\RelSpec}{\operatorname{\text{\ul{$\mathrm{Spec}$}}}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}

@@

@@

@@

@@

  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& R \\
    & A' \ar@{-->}[rru]
  } \]

@@

@@

@@

  \hil{Solution:} Define internally the frame of~$\RelSpec_X \A$ to be the frame of
  those radical ideals~$I \subseteq \A$ such that
  \[ \forall f\?\O_X\_ \forall s\?\A\_
    (\text{$f$ invertible in~$\O_X$} \Rightarrow s \in I) \Longrightarrow
      fs \in I. \]
  \pause
  Its \hil{points} are those prime filters~$G$ of~$\A$ such that
  \[ \forall f\?\O_X\_ \varphi(f) \in G \Rightarrow \text{$f$ invertible in~$\O_X$}. \]
\end{frame}

@@

@@

@@

@@

%  \begin{itemize}
%    \item This is a highbrow way of saying ``rational functions are regular
%    functions which are defined on a dense open subset''.
%    \item There is a generalization to nonreduced schemes.
%  \end{itemize}
\end{frame}

@@

  Observe that $M^{\sim} \cong \underline{M}[F^{-1}]$ is the localization of~$M$ at
  the \hil{generic filter}. Therefore:
  \begin{center}
    $M^{\sim}$ inherits all those properties of~$M$ which are \\
    \hil{stable under localization}.
  \end{center}
  Examples: finitely generated, free, flat, \ldots

@@

@@

@@

  \begin{itemize}
    \item \ldots{} $X$-schemes look just like sets,
    \item \ldots{} $\mathbb{P}^n_X$ is given by the naive expression
    \[ \{ (x_0,\ldots,x_n) \,|\, x_1 \neq 0 \vee \cdots \vee x_n \neq 0
    \}/\text{(rescaling)}, \]
    \item \ldots{} the cotangent ``bundle'' of an~$X$-scheme~$T$ is
    \[ \text{the set of maps $\Delta \to T$,} \]
    where~$\Delta = \{ \varepsilon \in \mathbb{A}^1_X \,|\, \varepsilon^2 = 0
    \}$.
    \item \ldots{} affinity is a ``double dual condition'', and
    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ s.\,th.
    \[ \forall f : \mathbb{A}^1_X[T]\_\
      \text{$f$ is monic separable} \Rightarrow
      \Diamond(\exists t : \mathbb{A}^1\_ f(t) = 0). \]
  \end{itemize}
\end{frame}

\note{\justifying
  \begin{itemize}
    \item The functor of points of~$\mathbb{A}^1_X$, that is
    \[ (T/X) \longmapsto \O_T(T), \]
    looks like a local ring and indeed like a field from the internal point of
    view, in the sense that
    \[ \forall f\?\mathbb{A}^1_X\_ \neg(f = 0) \Rightarrow \text{$f$
    invertible}. \]
    \item Describing the functor of points of the projective space was
    suggested by Zhen Lin Low.
    \item The statement on the étale topology follows from Gavin Wraith's
    article \emph{Generic Galois theory of local rings}.
  \end{itemize}
  \par
}

\begin{frame}\frametitle{Translating internal statements}
  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
  morphism of sheaves on~$X$. Then:
  \allowdisplaybreaks
  \begin{align*}
    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
    \Longleftrightarrow\
    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
    &\qquad
      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
    &\qquad\qquad
        \text{for all open~$W \subseteq V$:} \\
    &\qquad\qquad\qquad
          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
    &\qquad
          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}
\end{frame}

@@

\begin{frame}\frametitle{Limits of locally ringed locales}
  Let~$(X_i)_i$ be a diagram of locally ringed locales.

  Its limit is given by \hil{relocalizing} the corresponding limit
  in~$\mathrm{RL}$.
\end{frame}
% Mention only orally?
% Phrase more categorically.
% Extend to toposes.

>>>>

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{mathtools}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}

@@

\newcommand{\A}{\mathcal{A}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Hom}{\mathrm{Hom}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\RelSpec}{\operatorname{\text{\ul{$\mathrm{Spec}$}}}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}

@@

\note{\justifying
  See the \href{https://github.com/iblech/internal-methods/raw/master/notes.pdf}{notes} for more dictionary entries.

  The simple definition of~$\K_X$ allows to give an internal account of the
  basics of the theory of Cartier divisors, for instance giving an easy
  description of the line bundle associated to a Cartier divisor.
  \par
}

@@

\note{\justifying
  The~$\Diamond$-translation is a generalization of the \emph{double negation
  translation}, which is well-known in logic. The double negation translation
  has the following curious property: A formula~$\varphi$ admits a classical
  proof if and only if the translated formula~$\varphi^{\neg\neg}$ admits an
  intuitionistic proof.

  The~$\Diamond$-translation has been studied before (see for instance Aczel:
  \emph{The Russell--Prawitz modality}, and Escardó, Oliva: \emph{The Peirce
  translation and the double negation shift}), but to the best of my know\-ledge,
  this application -- expressing the internal language of
  subtoposes in the internal language of the ambient topos -- is new.
  \par
}

\note{\justifying
  For ease of exposition, assume that~$X$ is irreducible with generic
  point~$\xi$. Let~$\Diamond \defeqv \neg\neg$.

  Then~$\Sh(X) \models \Diamond \varphi$ means that~$\varphi$ holds on a dense
  open subset of~$X$, while~$\Sh(X) \models \varphi^\Diamond$ means
  that~$\varphi$ holds at the generic point (taking stalks of all involved
  sheaves).

  The question~``does~$\varphi^\Diamond$ imply~$\Diamond\varphi$?'' therefore
  means: Does~$\varphi$ spread from the generic point to a dense open subset?

  For the special case of the double negation translation, a general answer to
  this purely logical question has long been known: This holds if~$\varphi$ is
  a \emph{geometric formula} (doesn't contain~$\Rightarrow$ and~$\forall$).
  \par
}

\note{\justifying
  Let~$\F$ be a sheaf of modules on a locally ringed space~$X$.
  Assume that the stalk~$\F_x$ at some point~$x \in X$ vanishes.
  Then in general it does \emph{not} follow that~$\F$ vanishes on some open
  neighbourhood of~$x$.

  This can be understood in logical terms: The statement that~$\F$ vanishes,
  \[ \forall s : \F\_\ s = 0, \]
  is not a geometric formula.

  However, if~$\F$ is additionally supposed to be of finite type, then it
  \emph{does} follow that~$\F$ vanishes on an open neighbourhood. This too can
  be understood in logical terms: If~$\F$ is of finite type, then internally
  there are generators~$s_1,\ldots,s_n$ of~$\F$. Thus the vanishing of~$\F$ can
  be reformulated as
  \[ s_1 = 0 \wedge \cdots \wedge s_n = 0, \]
  and this condition is manifestly geometric.
  \par
}

@@

\note{\justifying
  The sheaf condition and the sheafification functor can be described purely
  internally. An object~$M$ is \emph{separated} with respect to~$\Diamond$ if
  and only if, from the internal point of view,
  \[ \forall x,y : M\_\ \Diamond(x = y) \Rightarrow x = y. \]
  It is a \emph{sheaf} with respect to~$\Diamond$, if furthermore
  \[ \forall K \subseteq M\_\
    \Diamond(\exists x : M\_\ K = \{ x \}) \Longrightarrow
    \exists x : M\_\ \Diamond(x \in K). \]

  The second condition displayed on the previous slide is equivalent to the
  separatedness condition. In the special case~$\E = \O_X$, $s = 1$ it reduces
  to Mulvey's ``somewhat obscure formula''. We now understand this condition in
  its proper context.
  \par
}

@@

  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

@@

\note{\justifying
  Details on this point of view can be found in one of Peter Arndt's very nice
  answers on MathOverflow:
  \begin{center}\url{http://mathoverflow.net/a/14334/31233}\end{center}
  \par
}

@@

\note{\justifying
  Monique Hakim constructed in her thesis a very general spectrum functor,
  taking a ringed topos to a locally ringed one, using explicit calculations
  with sites.

  Using the internal language allows to reduce these calculations to a minimum.
  One constructs the spectrum as the sheaf topos over an internal
  locale and then uses the general theorem that toposes over the base~$\E$ are
  the same as toposes internal to~$\E$.

  As a byproduct one obtains that Hakim's spectrum is \emph{localic} over the base.
  \par
}

@@

  \hil{Solution:} Define internally the frame of~$\RelSpec_X \A$ to be the frame of
  those radical ideals~$I \subseteq \A$ such that
  \[ \forall f\?\O_X\_ \forall s\?\A\_
    (\text{$f$ invertible in~$\O_X$} \Rightarrow s \in I) \Longrightarrow
      fs \in I. \]
  \pause
  Its \hil{points} are those prime filters~$G$ of~$\A$ such that
  \[ \forall f\?\O_X\_ \varphi(f) \in G \Longrightarrow \text{$f$ invertible in~$\O_X$}. \]
\end{frame}

\note{\justifying
  The stated condition on~$I$ is, under the assumption that~$\A$ is
  quasicoherent, equivalent to the condition that~$I$ is quasicoherent (as
  an~$\O_X$-module).

  The relative spectrum is thus constructed as a certain sublocale of the
  absolute one. The two constructions coincide if and only if the dimension of
  the base scheme is~$\leq 0$.
  \par
}

@@

\note{\justifying
  One might wonder whether the absolute spectrum or the relative one is ``more
  fundamental''. The absolute spectrum can be expressed using the relative one,
  since
  \[ \Spec A = \RelSpec_{\Spec \ZZ} A^{\sim}, \]
  but the other way is not in general possible: The absolute spectrum is always
  (quasi-)compact, while the relative one is not in general.
  \par
}
% XXX: Give details.

@@

\appendix

\section{Spreading from points to neighbourhoods}

@@

\section{The meromorphic functions revisited}

@@

  \begin{itemize}
    \item This is a highbrow way of saying ``rational functions are regular
    functions which are defined on a dense open subset''.
    \item Another reformulation is that~$\K_X$ is the sheafification of~$\O_X$
    with respect to the~$\neg\neg$-modality.
    \item There is a generalization to nonreduced schemes.
  \end{itemize}
\end{frame}

\section{Transfer principles}

@@

  Observe that $M^{\sim} \cong \underline{M}[\F^{-1}]$ is the localization of~$M$ at
  the \hil{generic prime filter} and that~$M$ and the constant sheaf of
  modules~$\underline{M}$ share all first-order properties. Therefore:
  \begin{center}
    $M^{\sim}$ inherits all those properties of~$M$ which are \\
    \hil{stable under localization}.
  \end{center}
  Examples: finitely generated, free, flat, \ldots

@@

\section{Applications in algebra}

@@

\begin{frame}\frametitle{Hilbert's program in algebra}
  \scriptsize\justifying
  There is a way to combine some of the powerful tools of classical ring theory
  with the advantages that constructive reasoning provides, for instance
  exhibiting explicit witnesses. Namely we can devise
  a language in which we can usefully talk about prime ideals, but which
  substitutes non-constructive arguments by constructive arguments ``behind
  the scenes''. The key idea is to substitute the phrase ``for all prime ideals''
  (or equivalently ``for all prime filters'') by ``for the generic prime filter''.

  More specifically, simply interpret a given proof using prime filters
  in~$\Sh(\Spec A)$ and let it refer to~$\F \hookrightarrow \underline{A}$.

  \hspace*{-0.75cm}%
  \begin{tabular}{lll}
    \toprule
    Statement & constructive substitution & meaning \\\midrule
    $x \in \ppp$ for all~$\ppp$. &
    $x \not\in \F$. &
    $x$ is nilpotent. \\
    $x \in \ppp$ for all~$\ppp$ such that~$y \in \ppp$. &
    $x \in \F \Rightarrow y \in \F$. &
    $x \in \sqrt{(y)}$. \\
    $x$ is regular in all stalks~$A_\ppp$. &
    $x$ is regular in~$\ull{A}[\F^{-1}]$. &
    $x$ is regular in~$A$. \\
    The stalks~$A_\ppp$ are reduced. &
    $\ull{A}[\F^{-1}]$ is reduced. &
    $A$ is reduced. \\
    The stalks~$M_\ppp$ vanish. &
    $\ull{M}[\F^{-1}] = 0$. &
    $M = 0$. \\
%   The stalks~$M_\ppp$ are fin.\@ gen.\@ over~$A_\ppp$. &
%   $\ull{M}[\F^{-1}]$ is fin.\@ gen.\@ over
%   $\ull{A}[\F^{-1}]$. &
%   $M$ is fin.\@ gen.\@ over~$A$. \\
    The stalks~$M_\ppp$ are flat over~$A_\ppp$. &
    $\ull{M}[\F^{-1}]$ is flat over~$\ull{A}[\F^{-1}]$. &
    $M$ is flat over~$A$. \\
    The maps~$M_\ppp \to N_\ppp$ are injective. &
    $\ull{M}[\F^{-1}] \to \ull{N}[\F^{-1}]$ is injective. &
    $M \to N$ is injective. \\
    The maps~$M_\ppp \to N_\ppp$ are surjective. &
    $\ull{M}[\F^{-1}] \to \ull{N}[\F^{-1}]$ is surjective. &
    $M \to N$ is surjective. \\
    \bottomrule
  \end{tabular}

  This is related (in a few cases equivalent) to the \emph{dynamical methods in
  algebra} explored by Coquand, Coste, Lombardi, Roy, and others. Their
  approach applies more generally.
  \par
\end{frame}

\section{The gros Zariski topos}

@@

  \begin{itemize}
    \item \ldots{} $X$-schemes look just like sets,
    \item \ldots{} $\mathbb{P}^n_X$ is given by the naive expression
    \[ \{ (x_0,\ldots,x_n) \,|\, x_1 \neq 0 \vee \cdots \vee x_n \neq 0
    \}/\text{(rescaling)}, \]
    \item \ldots{} the cotangent ``bundle'' of an~$X$-scheme~$T$ is
    \[ \text{the set of maps $\Delta \to \ull{T}$,} \]
    where~$\Delta = \{ \varepsilon \in \ull{\AA}^1_X \,|\, \varepsilon^2 = 0
    \}$.
    \item \ldots{} affinity is a ``double dual condition'', and
    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ s.\,th.
    \[ \forall f : \ull{\AA}^1_X[T]\_\
      \text{$f$ is monic separable} \Rightarrow
      \Diamond(\exists t : \ull{\AA}^1\_ f(t) = 0). \]
  \end{itemize}
\end{frame}

\note{
  \begin{itemize}\justifying
    \item The functor of points of~$\AA^1_X$, that is
    \[ \ull{\AA}^1_X : (T/X) \longmapsto \O_T(T), \]
    looks like a local ring and indeed like a field from the internal point of
    view, in the sense that
    \[ \forall f\?\ull{\AA}^1_X\_ \neg(f = 0) \Rightarrow \text{$f$
    invertible}. \]
    \item Let~$\A$ be a quasicoherent~$\O_X$-algebra. Let~$\E$ be the
    induced~$\ull{\AA}^1_X$-algebra given by~$\E(T \xrightarrow{\mu} X) \defeq
    (\mu^* \A)(T)$. Then the internal Hom set~$[\E,
    \ull{\AA}^1_X]_{\ull{\AA}^1_X}$ of~$\ull{\AA}^1_X$-algebra morphisms is the
    functor of points of~$\RelSpec_X(\A)$.
    \item Let~$\mu : T \to X$ be quasicompact and quasiseparated.
    Then~$\mu$ is affine iff, from the internal point of view, the map
    \[ \ull{T} \longrightarrow [ [\ull{T},\ull{\AA}^1_X]^\flat, \ull{\AA}^1_X ]_{\ull{\AA}^1_X},\ x \longmapsto
    \ull{\ \ }(x) \]
    into the ``double dual'' is bijective.
  \end{itemize}
}

\note{
  \begin{itemize}\justifying
    \item Describing the functor of points of the projective space was
    suggested by Zhen Lin Low.
    \item The statement on the étale topology follows from Gavin Wraith's
    article \emph{Generic Galois theory of local rings}.
  \end{itemize}
}

\section{Basics about the internal language}

\note{\justifying
  More generally, for an object~$U$ of a topos~$\E$, we define the meaning of
  \[ U \models \varphi \qquad\text{($\varphi$ holds on~$U$)}. \]
  Writing~``$\E \models \varphi$'' is then an abbreviation for~``$1 \models
  \varphi$'', where~``$1$'' denotes the terminal object of~$\E$.

  In addition to soundness with respect to intuitionistic logic, the internal
  language has the following two important properties:
  \begin{itemize}
    \item \hil{Monotonicity:} If~$p : V \to U$ is an arbitrary morphism and~$U
    \models \varphi$, then also~$V \models \varphi$.
    \item \hil{Locality:} If~$p : V \to U$ is an epimorphism and~$V \models
    \varphi$, then also~$U \models \varphi$.
  \end{itemize}
}

\note{\justifying
  In the special case that~$\E = \Sh(X)$ is the topos of sheaves on a
  topological space (or locale)~$X$, the rules of the Kripke--Joyal semantics
  look as follows. We tersely write~``$U \models \varphi$'' instead
  of~``$\Hom(\ull{\ \ }, U) \models \varphi$ for open subsets~$U \subseteq X$.
  \scriptsize
  \newcommand{\Ll}{:\Longleftrightarrow}
  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
    U \models \varphi \wedge \psi &\Ll&
      \text{$U \models \varphi$ and $U \models \psi$} \\
    U \models \varphi \vee \psi &\Ll&
      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
    U \models \varphi \Rightarrow \psi &\Ll&
      \text{for all open~$V \subseteq U$: }
    \text{$V \models \varphi$ implies $V \models \psi$} \\
    U \models \forall f \? \F\_ \varphi(f) &\Ll&
      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
      \varphi(f)$} \\
    U \models \exists f \? \F\_ \varphi(f) &\Ll&
      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
    $U_i \models \varphi(f_i)$}
  \end{array} \]
}

\begin{frame}\frametitle{Translating internal statements I}
  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
  morphism of sheaves on~$X$. Then:
  \allowdisplaybreaks
  \begin{align*}
    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
    \Longleftrightarrow\
    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
    &\qquad
      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
    &\qquad\qquad
        \text{for all open~$W \subseteq V$:} \\
    &\qquad\qquad\qquad
          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
    &\qquad
          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}
\end{frame}

\begin{frame}\frametitle{Translating internal statements II}
  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
  morphism of sheaves on~$X$. Then:
  \allowdisplaybreaks
  \begin{align*}
    & \Sh(X) \models \speak{$\alpha$ is surjective} \\[0.5em]
    \Longleftrightarrow\
    & \Sh(X) \models \forall t\?\G\_ \exists s\?\F\_ \alpha(s) = t \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $t \in \G(U)$:} \\
    &\qquad
      \text{there exists an open covering~$U = \textstyle\bigcup_i U_i$ and} \\
    &\qquad
      \text{sections~$s_i \in \F(U_i)$ such that:} \\
    &\qquad\qquad
        \alpha|_{U_i}(s_i) = t|_{U_i} \\[0.5em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is an epimorphism of sheaves}
  \end{align*}
\end{frame}

\begin{frame}\frametitle{Translating internal statements III}
  Let~$X$ be a topological space (or locale) and let~$s, t \in \F(X)$ be global
  sections of a sheaf~$\F$ on~$X$. Then:
  \allowdisplaybreaks
  \begin{align*}
    & \Sh(X) \models \neg\neg(s = t) \\[0.5em]
    \Longleftrightarrow\
    & \Sh(X) \models ((s = t) \Rightarrow \bot) \Rightarrow \bot \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$ such that} \\
    &\qquad
      \text{for all open~$V \subseteq U$ such that} \\
    &\qquad\qquad
      s|_V = t|_V, \\
    &\qquad
      \text{it holds that~$V = \emptyset$,} \\
    &
      \text{it holds that~$U = \emptyset$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{there exists a dense open set~$W \subseteq X$ such that $s|_W = t|_W$} \\
  \end{align*}
\end{frame}

@@
